New results in $\sfrac{k}{n}$ Power-Hours

### 2013-12-31 trunk/powerhour/paper/paper.tex
Revise paper, establishing BM7L is complete with two players, yay

% \pagestyle{empty}

@@

\maketitle

@@

This approach scales much better than enumeration and is efficient for
all sorts of generalizations (it works best when the dimensionality is
low---i.e., two players---and the expressiveness is high---i.e., many
cup states). Since we already know BM8L is universal, the remaining
open problem is BM7L, whose results are in
Figure~\ref{fig:samplebm7l}. This is tantalizingly close to a complete
matrix, with only 4 cells missing! It may be possible to solve these
two cases manually, though a proof of non-existence would be
difficult.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players7states.png}
\end{center}\vspace{-0.1in}
\caption{ 22.1 billion samples of legal two-player BM7L
  configurations. Fewer configurations were sampled than in
  Figure~\ref{fig:samplebm6l} because they take somewhat longer than
  6-state games to simulate, and a smaller proportion of random games
  are legal.
  %
  All but four of the $(60 + 1)^2 = 3721$ cells are attainable: 
  Permutations of $\langle 11, 53 \rangle$ and $\langle 49, 53 \rangle$
  remain elusive. Note that 53 drinks was also unattainable in a solo
  BM7L power-hour; this may in some sense be the ``hardest'' number of
  shots to drink in \bmsl.
}
\label{fig:samplebm7l}
\end{figure}

@@

Note that all of the two-dimensional figures resemble one another even
though they are fundamentally different (adding players, adding
states, adding random trials). This suggests that the combinatorial
problem (``what outcomes are reachable from finite state machines that
look kind of like this?'') has some geometric structure.

Some of the patterns are easy to explain. The top-left half of the
matrix is more populous than the bottom right, for example. This is
because we can bound the total number of drinks by $60 / c$, where $c$
is the number of cups active in the game (same as the number of cups
in the starting states). The top-left half is the region where this
sum is less than or equal to 30; in two player games, both players
must start with a cup in order to get an outcome in the bottom-right
half. We also see distinct clumps around 0, 15, 30, 45, 60; these
correspond to simple fractions (``drink every other time''; ``drink
three of four times'') of 60. This is intuitive because the expressive
power of BMML comes from the ability to form cycles of cup states and
drink on some fraction of them. Clumps are formed around these values
because of the possibility of preludes leading into the cycles
(Figure~\ref{fig:solo31}) that either drink (adding to the total) or
don't (subtracting from it). Minor clumps form as echoes between the
major ones, because a player may participate in two cycles of
different length (Figure~\ref{fig:dual45}).

@@

@@

\subsection{Known results}
\begin{enumerate}
  \item With one player, we have exact bounds on what is possible in
    the generalized case. With 8 cup states, a single player can
    drink 0--60 shots. Since each player can just play independently,
    this result extends to any number of players in BM8L. With fewer
    than 8 cup states, not every $k$ can be achieved alone.
  \item With two players in BM3L or BM4L, it is not possible for one
    of the players to drink every $k$ even if the other player helps her
    out.
  \item With two players, we know that BM5L does allow one player to
    drink any $k_1$ if the other player assists. In fact we can
    achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1 \}$.
    No other $k_2$ can be used universally, though of course many
    other combinations are possible
    (Figure~\ref{fig:composite2player}).
  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
    unlikely, given that random exploration plateaus with about 95.65\%
    of the grid filled.
  \item Open: Can BM7L achieve all $\langle k_1, k_2 \rangle$? Seems
    possible, but there are four (really two with symmetry) unsolved
    combinations.
  \item With three players in BM3L, one player can drink any number of
    shots if the other two players help.
\end{enumerate}

@@

\paragraph{Teetotaller: Someone can drink nothing.} When $\langle k_1, 
\ldots, k_i, \ldots, k_p \rangle$ is achievable in \bmsl, so is
$\langle k_1, \ldots, 0, \ldots, k_p \rangle$. This conjecture would
be trivial if not for the rule that requires us to drink the contents
of a full cup if we want to pass it in a different state. This
conjecture is true for all the graphics presented in this
paper;\footnote{Actually, it is not established for (only) 11 minutes
  in Figure~\ref{fig:powerthreehours}, but this is not a proper
  \bmsl\ game as it takes place over 180 minutes. There may very well
  be solutions for 11 anyway; this is just a sample.} we can see that
cells in the 0 column are always filled when some other cell in that
row is filled. If this conjecture fails, it probably fails for BM1L or
BM2L which have the least freedom per player.

@@

In this paper I present some new results in \kn\ Power-Hour theory, as
well as correct the historical record of some inebriated missteps. We
see that stochastic simulation, abstract interpration, and sampling
are powerful tools for solving combinatorial drinking problems. We
established some firm results for the classic game and some bounds for
generalizations, as well as informally looked at some visualizations
of its geometric structure. However, there are still several open
problems in this field that demand further study.

>>>>

\pagestyle{empty}

@@

\maketitle \thispagestyle{empty}

@@

This approach scales much better than enumeration and is efficient for
all sorts of generalizations (it works best when the dimensionality is
low---i.e., two players---and the expressiveness is high---i.e., many
cup states). Since we already know BM8L is universal, the remaining
open problem is BM7L, whose results are in
Figure~\ref{fig:samplebm7l}. Indeed, after more than 30 billion
samples the matrix is completely filled in; we have found an example
configuration that achieves every outcome. Some of these were
extremely rare, such as the solution for $\langle 11, 53 \rangle$
(Figure~\ref{fig:dual11-53}). In BM7L, two players can drink any
amount.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min2players7states.png}
\end{center}\vspace{-0.1in}
\caption{ 30.7 billion samples of legal two-player BM7L
  configurations. Fewer configurations were sampled than in
  Figure~\ref{fig:samplebm6l} because they take somewhat longer than
  6-state games to simulate, and a smaller proportion of random games
  are legal. Moreover, we stop after finding a solution for every
  cell, proving that BM7L is complete! The last cells found---an
  earlier version of this paper held these as open problems!---were
  permutations of $\langle 11, 53 \rangle$
  (Figure~\ref{fig:dual11-53}) and $\langle 49, 53 \rangle$. Note that
  53 drinks was also unattainable in a solo BM7L power-hour; this may
  in some sense be the ``hardest'' number of shots to drink in \bmsl.
}
\label{fig:samplebm7l}
\end{figure}

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.90 \linewidth]{dual11-53.pdf}
\end{center}\vspace{-0.1in}
\caption{ A solution for the elusive $\langle 11, 53 \rangle$ two-player
  BM7L power hour! This one was found after 30.7 billion outcomes sampled,
  making it the most rare. By the end of the game, the two players are
  just passing back and forth two cups in state 5. A long lead-in beginning
  on the left-player's state 2 spans 12 different configurations (the 6
  cup states for the two players) before entering the length-two cycle.
  Both cups travel along this lead-in, with one three steps ahead of the
  other. The right player drinks on every step until reaching the cycle
  (and then never again) for 11; the left player drinks during the cycle
  plus a little extra during the lead-in for 53. This configuration is
  quite flexible because the two players can make fine adjustments to
  their drink total by drinking or not drinking on the lead-in rules,
  which are executed just once or twice.}
\label{fig:dual11-53}
\end{figure}

@@

Note that all of the two-dimensional figures resemble one another even
though they are fundamentally different (adding players, adding
states, adding random trials). Even samples from BMML with three
players (3D projected to 2D), which is shown in Figure~\ref{fig:3d},
produces a similar pattern. This suggests that the combinatorial
problem (``what outcomes are reachable from finite state machines that
look kind of like this?'') has some geometric structure.

Some of the patterns are easy to explain. The top-left half of the
matrix is more populous than the bottom right, for example. This is
because we can bound the total number of drinks by $60 \times c$,
where $c$ is the number of cups active in the game (same as the number
of cups in the starting states). The top-left half is the region where
this sum is less than or equal to 60; in two player games, both
players must start with a cup in order to get an outcome in the
bottom-right half. We also see distinct clumps around 0, 15, 30, 45,
60; these correspond to simple fractions (``drink every other time'';
``drink three of four times'') of 60. This is intuitive because the
expressive power of BMML comes from the ability to form cycles of cup
states and drink on some fraction of them. Clumps are formed around
these values because of the possibility of preludes leading into the
cycles (Figures~\ref{fig:solo31}, \ref{fig:dual11-53}) that either
drink (adding to the total) or don't (subtracting from it). Minor
clumps form as echoes between the major ones, because a player may
participate in two cycles of different length
(Figure~\ref{fig:dual45}).

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.90 \linewidth]{sample60min3players3states.png}
\end{center}\vspace{-0.1in}
\caption{490 million samples of three-player BMML Power Hours. The
  cube is rotated $15\,^{\circ}$ along each axis, the top-left corner is
  $\langle 0, 0, 0 \rangle$, and the bottom right is $\langle 60, 60,
  60 \rangle$. }
\label{fig:3d}
\end{figure}

@@

\subsection{Known results}
\begin{enumerate}
  \item With one player, we have exact bounds on what is possible in
    the generalized case. With 8 cup states, a single player can
    drink 0--60 shots. Since each player can just play independently,
    this result extends to any number of players in BM8L. With fewer
    than 8 cup states, not every $k$ can be achieved alone.
  \item With two players in BM3L or BM4L, it is not possible for one
    of the players to drink every $k$ even if the other player helps her
    out.
  \item With two players, we know that BM5L does allow one player to
    drink any $k_1$ if the other player assists. In fact we can
    achieve any $\langle k_1, k_2 \rangle$ where $k_2 \in \{ 0, 1 \}$.
    No other $k_2$ can be used universally, though of course many
    other combinations are possible
    (Figure~\ref{fig:composite2player}). Many pairs $\langle k_1, k_2
    \rangle$ are known to be unattainable; this was established by
    exhaustively testing all possible configurations.
  \item Open: Can BM6L achieve all $\langle k_1, k_2 \rangle$? Seems
    unlikely, given that random exploration plateaus with about 95.65\%
    of the grid filled.
  \item BM7L can achieve all $\langle k_1, k_2 \rangle$. This was
    established by sampling random games until we found an example
    for every $\langle k_1, k_2 \rangle$.
  \item With three players in BM3L, one player can drink any number of
    shots if the other two players help.
\end{enumerate}

@@

\paragraph{Teetotaller: Someone can drink nothing.} When $\langle k_1, 
\ldots, k_i, \ldots, k_p \rangle$ is achievable in \bmsl, so is
$\langle k_1, \ldots, 0, \ldots, k_p \rangle$. This conjecture would
be trivial if not for the rule that requires us to drink the contents
of a full cup if we want to pass it in a different state. This
conjecture is true for all the graphics presented in this
paper;\footnote{Actually, it is not established for (only) 11 minutes
  in Figure~\ref{fig:powerthreehours}, but this is not a proper
  \bmsl\ game as it takes place over 180 minutes. There should be
  solutions for 11, like in Figure~\ref{fig:dual11-53}; this is just
  a sample.} we can see that
cells in the 0 column are always filled when some other cell in that
row is filled. If this conjecture fails, it probably fails for BM1L or
BM2L which have the least freedom per player.

@@

In this paper I presented some new results in \kn\ Power-Hour theory, as
well as correct the historical record of some inebriated missteps. We
saw that stochastic simulation, abstract interpration, and sampling
are powerful tools for solving combinatorial drinking problems. We
established some firm results for the classic game and some bounds for
generalizations, as well as informally looked at some visualizations
of its geometric structure. However, there are still several open
problems in this field that demand further study.
